---
title: "Gapminder"
author: "Cheryl"
date: "3/13/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```
## Load Dataset
Download the gapminder data frame from a website using tidyverse read_csv function
```{r}
gapminder <- read_csv("https://raw.githubusercontent.com/jules32/2017-11-30-MBARI/gh-pages/data/gapminder.csv")
```
Check out the first and last few lines of the data frame using head and tail
```{r}
head(gapminder)
tail(gapminder)
```
Look at the structure of the data frame
```{r}
str(gapminder)
```
Look at data frame dimensions, number of rows & columns
```{r}
dim(gapminder)
ncol(gapminder)
nrow(gapminder)
```
Look at the number of unique countries in the data frame using $ to select a column or
using the brackets (more specific)
```{r}
unique(gapminder$country)
unique(gapminder[,"continent"])
```
Get summary statistics of data frame
```{r}
summary(gapminder)
```
Install skim package
```{r}
#install.packages("skimr") # why use comments here?
```
Add skimr library and run skim summary stats function on gapminder data
```{r}
library("skimr") # attach library to current working environment
skim(gapminder)
```
Filter data using filter function
```{r}
filter(gapminder)
```
## Using dplyr
```{r}
dplyr::filter(gapminder, country == "Switzerland")  # run filter function in dplyr
```
Using the `dplyr::filter` format tells R to use the `dplyr` version of `filter`
```{r}
dplyr::filter(gapminder, lifeExp < 30)  # filter on life expectancy under 30
```

```{r}
dplyr::filter(gapminder, country == "Rwanda")  # filter on life expectancy under 30
```
Or check most recent year in Rwanda. How to do two countries at the same time? 
```{r}
filter(gapminder, country == "Rwanda", year == max(year))
```
Use select function in `dply` to select only certain columns
```{r}
dplyr::select(gapminder, country, year, pop) # select subset of columns from dataframe

```
Remove a column and make new data frame. Then `filter` using that data frame for the year 2007 only
```{r}
gapminder_no_cont <- dplyr::select(gapminder, -continent) # remove a single column from dataframe
dplyr::filter(gapminder_no_cont, year == 2007) 
```
Use pipe operator `(%>%)` to avoid creating temporary files on same operation above
```{r}
dplyr::select(gapminder, -continent) %>% 
  filter(year == 2007) 
```
Use `mutate` to create a new column in the data frame using math and a pipe to then sort the data in descending order
```{r}
gapminder %>%
  mutate(gdp = gdpPercap * pop) %>%
  arrange(desc(gdp))
```

Group data by country, `mutate` data frame and `ungroup`
```{r}
gapminder %>%  # data frame to work on
  group_by(country) %>%  # pull out country column
  mutate(gdp = gdpPercap * pop, max_gdp = max(gdp)) %>%  # create 2 new columns using mutate 
  ungroup()
```
```{r}
gapminder %>%
  group_by(country) %>%
  mutate(gdp = gdpPercap * pop, max_gdp = max(gdp)) %>%  # create 2 new columns using mutate 
  summarize(max_gdp = max(gdp)) %>% # summarize and take the max
  ungroup() # always ungroup to be sure you have ended the group_by
```
## Using Join
Now we are going to join or combine two tables together using a unique ID in dplyr
options are: left_join, right_joint, inner-Join, full_join
see https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf

Read in CO2 data for a subset of countries. It's good practice to name variables when reading in new data frames (avoid caps, be consistent)
```{r}
co2 <- readr::read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv") 
```

https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv
```{r}
head(co2)
summary(co2)
```
Now lets join this CO2 data with the gapminder dataset
```{r}
gap_2007 <- gapminder %>%
  filter(year==2007)  #filter on year
```
View gap_2007 data and check length of country list
```{r}
# View(gap_2007)
length(unique(gapminder$country)) # find length of unique country list
```
### Try Join
Try a join by ID of "country"
```{r}
left_join(gap_2007, co2, by = "country") # 1st dataset is left, 2nd is right
right_join(gap_2007, co2, by = "country") # try right join
full_join(gap_2007, co2, by = "country") # try full join
```
Can use set_diff or anti-join to check
```{r}
AirPassengers
```
 ## Reshaping Data
 Use tidyr to tidy data by reshaping and summarizing (gather, spread, separate, unite)
 Note this is not the best example  
```{r}
tidyr::gather(as.data.frame(AirPassengers), month, passenger)
```
```{r}
gap_wide <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv")
```
 
